---
title: "Loading and organizing data in memory"
author: "Cherki Amine, Florent Chuffart"
date: "`r Sys.Date()`"
output: html_document
---

```{r, echo=FALSE, eval=TRUE,warning=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, results="hide")
if (!exists("mread.table")) mread.table = memoise::memoise(read.table)
```

Since `R` widely works in memory space, this vignette loads and organises data in the memory.

Memory cache mechanism is mainly based on: 

  - memoisation using `memoise` package [1]
  - testing if variable name exists in environment



# Prerequisits

Customs packages:

- https://github.com/fchuffar/epimedtools (for a verbose apply)

# Defining global parameters

We define following global parameters : 

- `tcga_project` the targeted TCGA cancer study name.
- `gene_symbol` the gene symbol of the gene that we want to study.

```{r echo=TRUE}
tcga_project="TCGA-LUSC"
# gene_symbol =  "H1F0"
# gene_symbol = "AKR1B10"
gene_symbol = "AKR1E2"
```
  
# Load studies

We load data omics data for the TCGA project `r tcga_project` [2] as well.

  - `s_cnv  ` the *copy number variation* study
  - `s_meth ` the *methylome*  study
  - `s_trscr` the *transcriptome* study

These variables will not ne directly manipulate, they need to be preprocessed according to gene promoter definitions.

```{r label="loading studies", echo=TRUE}
if (!exists("mreadRDS")) mreadRDS = memoise::memoise(readRDS)
s_cnv   = mreadRDS(paste0("tcga_studies/study_", tcga_project, "_cnv.rds"))
s_meth  = mreadRDS(paste0("tcga_studies/study_", tcga_project, "_meth.rds"))
s_trscr = mreadRDS(paste0("tcga_studies/study_", tcga_project, "_trscr.rds"))
```

# Dynamicaly building of global gene/probe index

Since, many methylation probes could be associated to a given gene promoter  (see 
fig. 2 [Jones_nrg_2012](http://epimed.univ-grenoble-alpes.fr/downloads/florent/01_momik/doc/Jones_nrg_2012.pdf) and 
fig. 2 [Weber_medsci_2008](http://epimed.univ-grenoble-alpes.fr/downloads/florent/01_momik/doc/Weber_medsci_2008.pdf)),
we compute gene/probe index and store it in `gene_indexed_probes` global variable.

`gene_indexed_probes` is a list named by gene names and containing vector of probe names.

```{r}
genes = mreadRDS("genes/bed_grch38_epimeddb.rds")
genes = genes[rownames(genes)%in%rownames(s_trscr$data),]
if (!exists("gene_indexed_probes")) {
  # params
  pf_chr_colname = "Chromosome"
  pf_pos_colname = "Start"
  up_str = 2000
  dwn_str = 2000  
  if (!exists("pf_orig")) {
    pf_orig = s_meth$platform
    pf_orig = pf_orig[order(pf_orig[[pf_chr_colname]],pf_orig[[pf_pos_colname]]), ]
  }  
  ## index meth probes by chr
  chrs = unique(genes[,1])
  chrs_indexed_methpf = lapply(chrs, function(chr) {
    print(chr)
    idx = rownames(pf_orig)[!is.na(pf_orig[[pf_chr_colname]]) & pf_orig[[pf_chr_colname]]==chr]  
    ret = pf_orig[idx,]
    return(ret)
  })
  names(chrs_indexed_methpf) = chrs
  ## index probes by gene name
  print("# indexing probes by gene name")
  gene_indexed_probes = epimedtools::monitored_apply(genes, 1, function(gene) {
    # gene = randomall_genes[1,]genes=readRDS("~/fchuffar/projects/genes/bed_grch38_epimeddb.rds")
    # print(gene)
    chr = gene[[1]]
    meth_platform = chrs_indexed_methpf[[chr]]
    ret = dmprocr::get_probe_names(gene, meth_platform, pf_chr_colname, pf_pos_colname, up_str, dwn_str) 
    return(ret)
  })
  barplot(table(sapply(gene_indexed_probes, length)))
  gene_indexed_probes = gene_indexed_probes[sapply(gene_indexed_probes, length)>0]
}
```

```{r echo=TRUE, results=TRUE}
head(names(gene_indexed_probes))
length(gene_indexed_probes)
gene_indexed_probes[[gene_symbol]]
```

# The `mget_multiomic_data` function

We define a usefull function `mget_multiomic_data(gene, tcga_project)` that computes multiomic data of targeted TCGA cancer study for a given gene.

**Warning**, `mget_multiomic_data` uses `gene_indexed_probes` global variable without passing it as an argument of the function. This is imposed by the memoisation mechanism.

```{r label="mget_multiomic_data"}
if (!exists("mget_multiomic_data")) {
  # debugged and optimized version of dmprocr::trscr_meth_analysis https://github.com/bcm-uga/dmprocr
  preproc_omics_data = function(gene_symbol, s_cnv, s_meth, s_trscr, gene_indexed_probes) {
      meth_probe_idx = intersect(gene_indexed_probes[[gene_symbol]], rownames(s_meth$data))
      if (length(meth_probe_idx) <= 1) {
          return(NULL)
      }
      meth_data = s_meth$data[meth_probe_idx, ]
      meth_data = meth_data[, apply(is.na(meth_data), 2, sum)/nrow(meth_data) < 0.5]
      meth_data = meth_data[apply(is.na(meth_data), 1, sum)/ncol(meth_data) < 0.5, ]
      # dim(meth_data)
      meth_probe_idx = rownames(meth_data)
      if (!is.null(s_cnv)) {
          idx_sample = intersect (
            intersect(
              colnames(s_trscr$data)[order(s_trscr$data[gene_symbol,])], 
              colnames(meth_data)
            ),
            colnames(s_cnv$data)[abs(s_cnv$data[gene_symbol, ]) < 0.2]
          )
      } else {
          idx_sample = intersect(
            colnames(s_trscr$data)[order(s_trscr$data[gene_symbol,])], 
            colnames(meth_data)
          )
      }
      if (length(idx_sample) <= 1) {
          return(NULL)
      }
      meth_data = meth_data[, idx_sample]
      trscr_data = s_trscr$data[gene_symbol, idx_sample]
      ret = list(
        gene_symbol=gene_symbol, 
        study_name=s_trscr$stuffs$name, 
        meth_data=meth_data,
        trscr_data=trscr_data
      )
      if (!all(all(names(trscr_data) == colnames(meth_data)))) stop("Sample are not ordered in trscr_data like in meth_data")
      return(ret)
  }
  get_multiomic_data = function(gene_symbol, tcga_project) {
    # warning: gene_indexed_probes is a global variable
    s_cnv   = mreadRDS(paste0("tcga_studies/study_", tcga_project, "_cnv.rds"))
    s_meth  = mreadRDS(paste0("tcga_studies/study_", tcga_project, "_meth.rds"))
    s_trscr = mreadRDS(paste0("tcga_studies/study_", tcga_project, "_trscr.rds"))
    preproc_omics_data(gene_symbol, s_cnv, s_meth, s_trscr, gene_indexed_probes)
    
  }
  mget_multiomic_data = memoise::memoise(get_multiomic_data)
}
```




```{r echo=TRUE, results=TRUE}
data = mget_multiomic_data(gene_symbol, "TCGA-LUSC")
names(data)
data$gene_symbol
data$study_name
dim(data$meth_data)
head(data$meth_data[,1:3])
length(data$trscr_data)
head(data$trscr_data)
```











# Visualisation

```{r}
par(mar=c(10, 4.1, 4.1, 2.1))
layout(matrix(c(1, 1, 2, 2, 2, 2), 2), respect=TRUE)
# transcriptome
# par(mar=c(10, 4.1, 4.1, 2.1))
plot(data$trscr_data, 1:length(data$trscr_data), 
  main=paste(data$gene_symbol, tcga_project), 
  xlab="log2(normalized expression)", 
  ylab=paste0(length(data$trscr_data), " samples"), 
  yaxt="n"
) 
# methylome
colors = c("cyan", "black", "red")
cols = colorRampPalette(colors)(20)
breaks = seq(0, 1, length.out = length(cols) + 1)
main = paste0("methylome TSS+/-", up_str)
# par(mar=c(10, 4.1, 4.1, 2.1))
image(data$meth_data, col=cols,  breaks=breaks, xaxt="n", 
    yaxt="n", main=main)
axis(1, (1:nrow(data$meth_data) - 1)/(nrow(data$meth_data) - 1), rownames(data$meth_data), 
    las = 2)
```

# Scoring

A first score has been defined by considering the linear model 

$$expression~mean_metylation$$

where expression is explained by the mean of methylome over considered probes



```{r echo=TRUE, results=TRUE}
expression = data$trscr_data
mean_metylation = apply(t(data$meth_data),1,mean)
m = lm(expression~mean_metylation)
summary(m)
# linear regression
nb.pval <- length(rownames(data$meth_data))
pval <- rep(NA,nb.pval)
ml <- rep(NA,nb.pval)

for(i in 1 : nb.pval){
   ml <- lm(expression~data$meth_data[i,])
   pval[i] <- summary(ml)$coefficients[2,4]
}
names(pval) <- rownames(data$meth_data)
pval
```


```{r}
# construction d'un fichier bed
prob <- rownames(data$meth_data)
bed <- s_meth$platform[prob,1:3]
bed$gene_symbol <- data$gene_symbol
bed$pval <- pval
bed$strand <- (rep(genes$strand[genes$gene_symbol=="AKR1E2"],length(prob)))
```

**SGCE ET PEG10**

```{r}
data.sgce <- mget_multiomic_data("SGCE", "TCGA-LUSC")
data.peg10 <- mget_multiomic_data("PEG10", "TCGA-LUSC")
meth.sgce <- data.sgce$meth_data
meth.peg10 <- data.peg10$meth_data
trscr.sgce <- data.sgce$trscr_data
trscr.peg10 <- data.peg10$trscr_data

# suppression des NA pour SGCE
size.sgce <- length(data.peg10$trscr_data)
for(i in 1:size.sgce){
  for(j in 1:length(rownames(meth.sgce))){
    if(is.na(meth.sgce[j,i])==TRUE){
      meth.sgce[j,i] <- mean(meth.sgce[,i],na.rm = TRUE)
    }
  }
}
# suppression des NA pour PEG10
for(i in 1:length(data.peg10$trscr_data)){
  for(j in 1:length(rownames(meth.peg10))){
    if(is.na(meth.peg10[j,i])==TRUE){
      meth.peg10[j,i] <- mean(meth.peg10[,i],na.rm = TRUE)
    }
  }
}
which(is.na(meth.peg10))
which(is.na(meth.peg10))

# linear regression
pval1 <- c()
for(i in 1:length(rownames(meth.peg10))){
  lm2 <- lm(meth.sgce[i,]~trscr.sgce)
  pval1[i] <- summary(lm2)$coefficients[2,4]
}
```



```{r}
# construction d'un fichier bed
prob.sgce <- rownames(meth.sgce)
bed.sgce <- s_meth$platform[prob.sgce,1:3]
bed.sgce$gene_symbol <- data.sgce$gene_symbol
bed.sgce$pval <- pval1
bed.sgce$strand <- (rep(genes$strand[genes$gene_symbol=="SGCE"],length(prob.sgce)))
head(bed.sgce)
prefix = "sgce"
write.table(bed.sgce,file = paste0(prefix,".bed"),sep="\t",quote=FALSE,row.names = FALSE,col.names = TRUE)
```
```{r}
par(mfrow=c(1,2))
# les p-valeurs corrigées avec comb-p
plot(1:length(trscr.sgce),trscr.sgce,xlab = "233",ylab="valeurs d'expression")
# couleurs du niveau de la methylation
colors <- c("cyan", "black", "red")
cols <- colorRampPalette(colors)(20)
breaks <- seq(0, 1, length.out = length(cols) + 1)
main <- paste0("methylome TSS+/-", up_str)
image(meth.sgce,col = cols,breaks=breaks,main=main)
axis(1, (1:nrow(meth.sgce) - 1)/(nrow(meth.sgce) - 1), rownames(meth.sgce), 
    las = 2)
```

```{r}
comb.p <- read.table("data/pvalslk.txt")
pval.c <- comb.p$V5
plot(pval1,col="red",pch=19,main = "p-valeurs SLK vs p-valeurs de la regression")
points(pval.c,col="blue",pch=19)
legend("topleft",legend = c("p-valeur.reg","pval.slk"),col = c("red","blue"),pch=19)
```

```{r eval=FALSE}
layout(matrix(1:2,1), respect=TRUE)
foo = apply(data$meth_data, 1, function(meth){
  # meth = data$meth_data[1,]
  m = lm(data$trscr_data~meth)
  beta = m$coefficients[[2]]
  pval = anova(m)[1,5]
  c(beta,pval)
})

plot(foo[1,], -log10(foo[2,]), main="Volcano plot", xlab="beta", ylab="-log10(pv)")

names(data$trscr_data) == colnames(data$meth_data)
foo = apply(data$meth_data, 1, function(meth){
  m = MASS::rlm(data$trscr_data~meth, maxit=400)  
  beta = m$coefficients[[2]]
  pval = survey::regTermTest(m, "meth", null=NULL,df=Inf, method=c("Wald"))$p
  c(beta,pval)
})
plot(foo[1,], -log10(foo[2,]))

```


# References

[1] `memoise` package, https://cran.r-project.org/web/packages/memoise (april 2020).
[2] TCGA project LUSC, https://portal.gdc.cancer.gov/projects/TCGA-LUSC (april 2020).
[Jones_nrg_2012] http://epimed.univ-grenoble-alpes.fr/downloads/florent/01_momik/doc/Jones_nrg_2012.pdf
[Weber_medsci_2008] http://epimed.univ-grenoble-alpes.fr/downloads/florent/01_momik/doc/Weber_medsci_2008.pdf
[EPIC] chap. 2,pages 2-9, https://support.illumina.com/content/dam/illumina-support/documents/documentation/chemistry_documentation/infinium_assays/infinium_hd_methylation/infinium-hd-methylation-guide-15019519-01.pdf



# Session Information

```{r, results="verbatim"}
sessionInfo()
```




